Fail Faster

### 2025-03-16 paper/paper.tex
comments on S1

Property-based testing (PBT) uses generators of random data to produce 
inputs to a system under test, which is then checked against a set of 
properties the system is expected to satisfy.  In practice, PBT is used 
as a lightweight, ad hoc approach to software correctness, aimed at 
detecting bugs within a short testing window---typically 50 milliseconds 
to 30 seconds. Given this practical constraint, it is imperative that 
inputs be generated as fast as possible; in other words, generators must be 
treated as performance-sensitive code.

The locus for improvement in this area is \textit{generator libraries}: if 
programmers are provided with the tools to write fast generators, then they 
do not themselves need to be performance experts. However, existing generator 
libraries fail to achieve good performance due to two major sources of 
inefficiency. The first is \textit{random sampling}, or the process of 
selecting random values, which, depending on one's choice of \rand and
the number of samples needed, can constitute an unexpectedly high proportion
of a generator's runtime. The second is \textit{abstraction}: while flexible 
and expressive, generator libraries rely on dynamic control structures that
allocate at runtime. Allocations compound in the aggregate, leading to costly
garbage collection pauses. This raises a fundamental 
question: can generator libraries produce code that is both expressive 
\textit{and} efficient?

  % in strict functionallanguages like OCaml and Scala
We outline a fully general, two-stage approach to designing generator 
libraries so that they retain their idiomatic abstractions with 
significantly improved performance. First, we demonstrate the impact of
RNG choice by optimizing an existing \rand. Our optimized RNG produces 
exactly 
the same values as its unoptimized counterpart, allowing us to directly 
measure the impact of this choice on generator performance. Second, 
we propose a method of streamlining generator libraries using 
\textit{staged metaprogramming}, or ``staging''. \tr{Joe---maybe add
a sentence here about parsing literature?} Using staging, it is possible
to eliminate the overhead of many common generator abstractions. For example,
where an unstaged generator DSL might represent a choice between
a list of options as a dynamically-allocated array of closures, a staged 
generator can expand to an if-else tree at compile 
time---eliminating the allocation entirely. This approach is not tied to any
particular PBT library, but rather provides a recipe for improving even
state-of-the-art implementations.

To demonstrate the effectiveness of our technique, we apply it to a leading PBT 
library,\\ \bq, built with performance engineering in mind.
Our version of this library, termed \name, is a drop-in replacement for \bq which 
is completely semantically equivalent: that is,  generators written using \name 
produce exactly the same values as those written in \bq. However, in \name, 
the abstractions provided by the generator DSL are zero-cost, and the \rand is 
highly-optimized. The direct equivalence between \name and \bq allows us to 
perform extremely fine-grained comparisons between generators, as generators 
implemented using both versions of the library will produce exactly the same 
sequence of values; so, differences in testing performance are directly 
attributable to \name's performance speedup.
\tr{Not really done... }

To test \name's effectiveness, we run a bunch of experiments to show that \name runs faster, allocates less, etc., than
\bq on a variety of different workloads. We demonstrate the generality of the technique by implementing \name in
Scala, and benchmarking the same generators in the Scala port of the library. Finally, as a sanity check, we demonstrate
that \name allows you to find bugs significantly faster ...

>>>>

Property-based testing (PBT) uses generators of random data to produce 
inputs to a system under test, which is then checked against a set of 
properties the system is expected to satisfy.  In practice, PBT is used 
as a lightweight, ad hoc approach to software correctness,
\jwc{Move this bit about how it's used to the first sentence. I.e. PBT is a testing technqique where random data is fed to a system under test...}
aimed at 
detecting bugs within a short testing window---typically 50 milliseconds 
to 30 seconds. \jwc{This sells us a little short: we should say something to the effect of ``Recent work on understanding how PBT is used in practice has determined that practitioners often run PBT with very short time budgets...''} Given this practical constraint, it is imperative that 
inputs be generated as fast as possible; in other words, generators must be 
treated as performance-sensitive code.

The locus for improvement in this area \jwc{``This area'' sounds like it's talking about the whole area of PBT... Maybe ``a high-leverage way to improve the performance of generators is to...''} is \textit{generator libraries}: if 
programmers are provided with the tools to write fast generators, then they 
do not themselves need to be performance experts. However, existing generator 
libraries fail to achieve good performance \jwc{sounds too absolute --- who's to say what ``good performance'' is?} due to two major sources of 
inefficiency. The first \jwc{swap these? we talk about them in the other order} is \textit{random sampling}, or the process of 
selecting random \proposechange{values}{numbers} \jwc{i'd suggest we use ``values'' to mean BSTs and things}, which, depending on one's choice of \rand and
the number of samples needed, can constitute an unexpectedly high proportion
of a generator's runtime \jwc{Nice, like this}. The second is \textit{abstraction}: while flexible 
and expressive, generator libraries rely on dynamic control structures that
allocate at runtime. Allocations compound in the aggregate, leading to costly
garbage collection pauses. \jwc{Allocation cost is a slightly OCaml-specific problem compared to the more general issue that the abstractions (binds, etc) are \emph{opaque} to the compiler, so it can't optimize the code at all.} This raises a fundamental 
question: can generator libraries produce code that is both expressive 
\textit{and} efficient? \jwc{love this end to the paragraph.}

  % in strict functionallanguages like OCaml and Scala
We outline a \proposecut{fully general}, two-stage \jwc{two-pronged? this clashes with staging i.e. metaprogramming} approach to designing generator 
libraries so that they retain their idiomatic abstractions with 
significantly improved performance \jwc{simplify: idiomatic and performant}. First, we demonstrate the impact of
RNG choice by optimizing an existing \rand. Our optimized RNG produces 
exactly 
the same values as its unoptimized counterpart, allowing us to directly 
measure the impact of this choice on generator performance. \jwc{Permute? I think staging \emph{then} randomness.} \jwc{Also, i don't love this as the first place we explain the drop-in equivalence of the RNG. It's unmotivated why wen want this.} Second, 
we propose a method of streamlining \jwc{``streamlining feels too vague here''} generator libraries using 
\textit{staged metaprogramming}, or ``staging''. \tr{Joe---maybe add
a sentence here about parsing literature?} \jwc{Maybe something like: ``Staging has long been used across domains as a technique of eliminating the overhead of idiomatic functional DSLs (cite a bunch of shit).''} Using staging, it is possible
to eliminate the overhead of many common generator abstractions. For example,
\jwc{I think we should use return/bind as the simplest example, since then you don't have to get into the weeds of allocation or closures. The functional abstractions of a monadic generator DSL block the compiler from generating fast code.}
\jwc{Then present this case as ``more sophisticated use of staging'' or something?}
where an unstaged generator DSL might represent a choice between
a list of options as a dynamically-allocated array of closures, a staged 
generator can expand to an if-else tree at compile 
time---eliminating the allocation entirely. This approach is not tied to any
particular PBT library, but rather provides a recipe for improving even
state-of-the-art implementations \jwc{Sentence structure doesn't follow. Instead maybe ``not any particular, but a recipe for improving in general''}.

To demonstrate the effectiveness of our technique, we apply it to a leading PBT 
library, \bq, built with performance engineering in mind. \jwc{Unclear if this is true, I think we should just say that by our testing it's the fastest in OCaml.}
Our version of this library, termed \name, is a drop-in replacement for \bq which 
is completely semantically equivalent: that is, \jwc{given identical random seeds} generators written using \name 
produce exactly the same values as those written in \bq. However, in \name, 
the abstractions provided by the generator DSL are zero-cost, and the \rand is 
highly-optimized. The direct equivalence between \name and \bq allows us to 
perform extremely fine-grained comparisons between generators, as generators 
implemented using both versions of the library will produce exactly the same 
sequence of values; so, differences in testing performance are directly 
attributable to \name's performance speedup.
\tr{Not really done... }
\jwc{The preceeding paragraph is \emph{really good}}

To test \name's effectiveness, we run a bunch of experiments to show that \name runs faster, allocates less, etc., than
\bq on a variety of different workloads. \jwc{Move this to previous para. I think this para should start with ``To demonstrate the generality'', paralleling the ``to demonstrate the effectiveness''}We demonstrate the generality of the technique by implementing \name in
Scala, and benchmarking the same generators in the Scala port of the library. Finally, as a sanity check, we demonstrate
that \name allows you to find bugs significantly faster ...
